# Random Variables

> Learning Objectives
>
>
> 1. Distinguish between **discrete** and **continuous** random variables  
> 2. Define random variables for real-world data problems  
> 3. Identify appropriate probability distributions for common data types  
> 4. Connect random variables to **data columns** used in SAS programs to represent a real world question


## What Is a Random Variable?

A **random variable (RV)** is a numerical quantity whose value depends on the outcome of a random experiment.

We typically denote a random variable by an uppercase letter, such as  $X$ and its realized value by a lowercase letter, such as  $X = x$. The random variable can be *continuous* or *discrete*.



In practice, we often observe multiple realizations, or running the random experiment multiple times, say $n$ times or $n$ realizations. We denote these realizations as:
$$X_1 = x_1, \; X_2 = x_2, \; \ldots, \; X_n = x_n.$$
The number $n$ is called the **sample size**.

::: {.callout-example }
Exercise: Define whether the following random variables are discrete or continuous, and the possible values that $X$ takes.

+ Number of emails received by a server in one hour
($X = 0, 1, 2, \ldots$: discrete)
+ Time (in minutes) until a machine fails
($X = x \in [0, \infty)$: continuous)
+ Total number of defects on a manufactured item
($X = 0, 1, 2, \ldots$: discrete)
+ Daily maximum temperature in Atlanta (in degrees Fahrenheit)
($X = x \in (-\infty, \infty)$: continuous)
+ Whether a randomly selected loan defaults within one year
($X = 1 \text{ if default, } 0 \text{ otherwise}$: discrete)
:::

::: {.callout-exercise}
Exercise: Define whether the following random variables are discrete or continuous, and the possible values that $X$ takes.

1. Number of transactions made by a customer in a day 
2. Response time (in seconds) of a web service request 
3. Count of hospital admissions in a city per week  discrete)
4. Proportion of time a system is idle during a day
:::

## Discrete vs Continuous Random Variables

### Discrete Random Variables

A **discrete random variable** takes values in a **finite or countable set**.

**Examples:**

- Number of heads in three coin flips  
- Number of students passing an exam  
- Number of events occurring in a fixed time period  

A discrete random variable is described by a **probability mass function (PMF)**:

| Value of $X$ | $x_1$ | $x_2$ | $x_3$ | … | $x_m$ |
|--------------|--------|--------|--------|---|--------|
| Probability | $p_1$ | $p_2$ | $p_3$ | … | $p_m$ |

These probabilities satisfy:

- $0 \le p_i \le 1$,
- $\sum_{i=1}^m p_i = 1$.

We calculate the probability of events modelled by discrete random variables by summing up the probability $p_i$ for the values $x_i$ that make up the event.

::: {.callout-exercise title = "Discrete Probability Calculation"}
Suppose the length $X$ (in minutes) of an international phone call has distribution:

| X | 1 | 2 | 3 | 4 |
|---|---|---|---|---|
| P(X) | 0.2 | 0.5 | 0.2 | 0.1 |

Then, calculate the following probabilities

a. $P(X \le 2)$
b. $P(X < 2)$
c. $P(X > 1)$
:::

## Common Discrete Distributions

### Bernoulli Distribution

Used for **binary outcomes**:

- success / failure
- yes / no
- 1 / 0

Notation: $X \sim \text{Ber}(p)$, where $p$ is the probability of success.

**Example:**  
Whether a patient has diabetes (1 = yes, 0 = no).

How to specify the probability $p$ will be discussed in the following lecture. This is related to the procedure of statistical inference.

### Poisson Distribution

Used to model **counts of events over time or space**.

Notation:
$X \sim \text{Poi}(\lambda)$

where $\lambda$ is the **mean rate**.

**Examples:**

+ Number of trades per day
+ Number of system failures per week
+ Number of arrivals to a service queue

Similar to the probability $p$ from the Bernoulli distribution, how to specify the rate $\lambda$ will be discussed in the following lecture.

## Continuous Random Variables

A **continuous random variable** can take **any value $x$ in an interval**.

**Examples:**

- Height of individuals
- Time until failure of a component
- Test scores treated as continuous

Probabilities are defined using **density functions**, not point probabilities. Some common continuous distributions are as follows.

### Normal Distribution

The normal distribution is also called the **Gaussian distribution**. It is characterized by two parameters: the mean $\mu$ and the standard deviation $\sigma$. It is unimodal and symmetric around the mean which is the centre of the mass. A continuous random variable $X$ that has a normal distribution is said to be *normal* or *normally distributed*. 

Notation: $X \sim N(\mu, \sigma^2)$. Sometimes the standard deviation may be used instead of the variance, which is the square of the variance.

Institution of the normal distribution:

- mean: $\mu$
- variance: $\sigma^2$

``` {r normal}
# Example: plot normal distribution with mean = 0, sd = 1
x <- seq(-4, 4, length=100)
y <- dnorm(x, mean=0, sd=1)
library(ggplot2)
# use ggplot2
data.frame(x, y) |>
  ggplot(aes(x=x, y=y)) +
  geom_line() +
  ggtitle("Normal Distribution N(0,1)") +
  xlab("x") + ylab("Density") + theme_minimal() +
  geom_vline(xintercept=0, linetype="dashed", color="red")
```

A distribution plot of the normal distribution with mean 0 and standard deviation 1 is shown above. It is often referred as the *standard normal distribution*. In practice, we would like to "standardized" the data to have mean 0 and standard deviation 1 for the subsequent analysis.

Normal distribution play an important role in statistical inference. One of the important property is the 68-95-99.7 rule:
- About 68% of the data falls within one standard deviation of the mean
- About 95% of the data falls within two standard deviations of the mean
- About 99.7% of the data falls within three standard deviations of the mean

``` {r 68-95-99.7}
x <- seq(-4, 4, length = 100)
y <- dnorm(x)

library(ggplot2)

df <- data.frame(x, y)

ggplot(df, aes(x = x, y = y)) +
  geom_line(linewidth = 1) +
  theme_minimal() +
  labs(
    title = "Normal Distribution with 68–95–99.7 Rule",
    x = NULL,
    y = "Density"
  ) +
  
  # remove numeric x-axis ticks and labels
  scale_x_continuous(breaks = NULL) +
  
  # vertical reference lines
  geom_vline(xintercept = 0, linewidth = 0.8) +
  geom_vline(xintercept = c(-1, 1), linetype = "dashed", color = "red") +
  geom_vline(xintercept = c(-2, 2), linetype = "dashed", color = "blue") +
  geom_vline(xintercept = c(-3, 3), linetype = "dashed", color = "darkgreen") +
  
  # percentage labels
  annotate("text", x = 0, y = 0.38, label = "68%", color = "red") +
  annotate("text", x = 0, y = 0.33, label = "95%", color = "blue") +
  annotate("text", x = 0, y = 0.28, label = "99.7%", color = "darkgreen") +
  
  # mu and sigma labels only
  annotate("text", x =  0, y = -0.01, label = expression(mu), vjust = 1.5) +
  annotate("text", x =  1, y = -0.01, label = expression(sigma),  vjust = 1.5, color = "red") +
  annotate("text", x = -1, y = -0.01, label = expression(-sigma), vjust = 1.5, color = "red") +
  annotate("text", x =  2, y = -0.01, label = expression(2*sigma),  vjust = 1.5, color = "blue") +
  annotate("text", x = -2, y = -0.01, label = expression(-2*sigma), vjust = 1.5, color = "blue") +
  annotate("text", x =  3, y = -0.01, label = expression(3*sigma),  vjust = 1.5, color = "darkgreen") +
  annotate("text", x = -3, y = -0.01, label = expression(-3*sigma), vjust = 1.5, color = "darkgreen")
```

### Exponential Distribution

The exponential distribution is another common distribution where a few outcomes are most likely with a rapidly decreasing probability for larger values. It is often used to model waiting times or lifetimes of objects. It is similar to the *geometric distribution* in the discrete case.

Two ways to specify the parameter.

1. $X \sim \text{Exp}(\lambda)$, where $\lambda$ is the *rate* parameter.
2. $X \sim \text{Exp}(\beta)$, where $\beta$ is the *scale* parameter, and $\beta = 1/\lambda$.

The notation is either $X \sim \text{Exp}(\lambda)$ or $X \sim \text{Exp}(\beta)$. But to be sure which parameterization is used, we need to check the definition of the distribution.

::: {.callout-example}
Please define adequate random variables for the following data example, and thaink about what distribution can be used to model the data.

+ Suppose we want to know whether the rate of diabetes of a certain area is too high. The researchers randomly discuss with 10 individuals in a certain area to know whether they have diabetes. The data are collected as "yes" or "no" answers as follows

| Individual | 1 | 2 | 3 | 4 | 5 | 6 | 7 | 8 | 9 | 10 |
|------------|---|---|---|---|---|---|---|---|---|----|
| Outcome    | 1 | 0 | 0 | 0 | 0 | 1 | 0 | 0 | 0 | 0  |

+ A comapny that manufactures light blulb claims that a particular type of light bulb will last 850 hours on average. To justify the claim more scientifically, a researcher randomly selects 10 light bulbs of that type and measures the lifetimes (in hours) of the light bulbs as follows:

| Light Bulb | 1   | 2   | 3   | 4   | 5   | 6   | 7   | 8   | 9   | 10  |
|------------|-----|-----|-----|-----|-----|-----|-----|-----|
| | 831 | 832 | 840 | 819 | 822 | 836 | 829 | 817 | 899 | 823 | 

:::

In the following classes we will assume the dataset we have can be well-modelled represented for the underline studies. The data collection, however, is beyond the scope of the class. For those interested in data collection, please refer to the *survey sampling* or *experimental design* area.


## Solution to the exercise

Answer of the exercise 

Exercise 1

1. Number of transactions made by a customer in a day ($X = 0, 1, 2, \ldots$: discrete)
2. Response time (in seconds) of a web service request ($X = x > 0$: continuous)
3. Count of hospital admissions in a city per week ($X = 0, 1, 2, \ldots$: discrete)
4. Proportion of time a system is idle during a day ($X = x \in [0,1]$: continuous)

Exercise 2:

- $P(X \le 2) = 0.7$
- $P(X < 2) = 0.2$
- $P(X > 1) = 0.8$